---
title: "Package MKinfer"
author: "Matthias Kohl"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
bibliography: MKinfer.bib
vignette: >
  %\VignetteIndexEntry{MKinfer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{utf8}
---


## Introduction  
Package MKinfer includes a collection of functions for the computation of 
various confidence intervals [@Altman2000; @Hedderich2018] including bootstrapped 
versions [@Davison1997] as well as Hsu [@Hedderich2018], permutation [@Janssen1997], 
bootstrap [@Davison1997] and multiple imputation [@Barnard1999] t-test.

We first load the package.
```{r}
library(MKinfer)
```


## Confidence Intervals
### Binomial Proportion
There are several functions for computing confidence intervals. We can compute
12 different confidence intervals for binomial proportions [@DasGupta2001]; e.g.
```{r}
## default: "wilson"
binomCI(x = 12, n = 50)
## Clopper-Pearson interval
binomCI(x = 12, n = 50, method = "clopper-pearson")
## identical to 
binom.test(x = 12, n = 50)$conf.int
```

For all intervals implemented see the help page of function binomCI. One can
also compute bootstrap intervals via function boot.ci of package boot [@Davison1997] 
as well as one-sided intervals.


### Difference of Two Binomial Proportions
There are several functions for computing confidence intervals. We can compute
different confidence intervals for the difference of two binomial proportions
(independent [@Newcombe1998a] and paired case [@Newcombe1998b]); e.g.
```{r}
## default: wilson 
binomDiffCI(a = 5, b = 0, c = 51, d = 29)
## default: wilson with continuity correction
binomDiffCI(a = 212, b = 144, c = 256, d = 707, paired = TRUE)
```

For all intervals implemented see the help page of function binomDiffCI. One
can also compute boostrap intervals via function boot.ci of package 
boot [@Davison1997]  as well as one-sided intervals.


### Mean and SD
We can compute confidence intervals for mean and SD [@Altman2000, @Davison1997].
```{r}
x <- rnorm(50, mean = 2, sd = 3)
## mean and SD unknown
normCI(x)
meanCI(x)
sdCI(x)
## SD known
normCI(x, sd = 3)
## mean known
normCI(x, mean = 2, alternative = "less")
## bootstrap
meanCI(x, boot = TRUE)
```


### Difference in Means
We can compute confidence interval for the difference of means [@Altman2000; 
@Hedderich2018; @Davison1997].
```{r}
x <- rnorm(20)
y <- rnorm(20, sd = 2)
## paired
normDiffCI(x, y, paired = TRUE)
## compare
normCI(x-y)
## bootstrap
normDiffCI(x, y, paired = TRUE, boot = TRUE)

## unpaired
y <- rnorm(10, mean = 1, sd = 2)
## classical
normDiffCI(x, y, method = "classical")
## Welch (default as in case of function t.test)
normDiffCI(x, y, method = "welch")
## Hsu
normDiffCI(x, y, method = "hsu")
## bootstrap: assuming equal variances
normDiffCI(x, y, method = "classical", boot = TRUE, bootci.type = "bca")
## bootstrap: assuming unequal variances
normDiffCI(x, y, method = "welch", boot = TRUE, bootci.type = "bca")
```


In case of unequal variances and unequal sample sizes per group the classical
confidence interval may have a bad coverage (too long or too short), as is 
indicated by the small Monte-Carlo simulation study below.
```{r}
M <- 100
CIhsu <- CIwelch <- CIclass <- matrix(NA, nrow = M, ncol = 2)
for(i in 1:M){
  x <- rnorm(10)
  y <- rnorm(30, sd = 0.1)
  CIclass[i,] <- normDiffCI(x, y, method = "classical")$conf.int
  CIwelch[i,] <- normDiffCI(x, y, method = "welch")$conf.int
  CIhsu[i,] <- normDiffCI(x, y, method = "hsu")$conf.int
}
## coverage probabilies
## classical
sum(CIclass[,1] < 0 & 0 < CIclass[,2])/M
## Welch
sum(CIwelch[,1] < 0 & 0 < CIwelch[,2])/M
## Hsu
sum(CIhsu[,1] < 0 & 0 < CIhsu[,2])/M
```


### Coefficient of Variation
We provide 12 different confidence intervals for the (classical) coefficient 
of variation [@Gulhar2012; @Davison1997]; e.g.
```{r}
x <- rnorm(100, mean = 10, sd = 2) # CV = 0.2
## default: "miller"
cvCI(x)
## Gulhar et al. (2012)
cvCI(x, method = "gulhar")
## bootstrap
cvCI(x, method = "boot")
```

For all intervals implemented see the help page of function cvCI.


### Quantiles, Median and MAD
We start with the computation of confidence intervals for quantiles [@Hedderich2018; @Davison1997].
```{r}
x <- rexp(100, rate = 0.5)
## exact
quantileCI(x = x, prob = 0.95)
## asymptotic
quantileCI(x = x, prob = 0.95, method = "asymptotic")
## boot
quantileCI(x = x, prob = 0.95, method = "boot")
```

Next, we consider the median.
```{r}
## exact
medianCI(x = x)
## asymptotic
medianCI(x = x, method = "asymptotic")
## boot
medianCI(x = x, method = "boot")
```

Finally, we take a look at MAD (median absolute deviation) where by default 
the standardized MAD is used (see function mad).
```{r}
## exact
madCI(x = x)
## aysymptotic
madCI(x = x, method = "asymptotic")
## boot
madCI(x = x, method = "boot")
## unstandardized
madCI(x = x, constant = 1)
```


## Hsu Two-Sample t-Test
The Hsu two-sample t-test is an alternative to the Welch two-sample t-test using
a different formula for computing the degrees of freedom of the respective 
t-distribution [@Hedderich2018]. The following code is taken and adapted from 
the help page of the t.test function.
```{r}
t.test(1:10, y = c(7:20))      # P = .00001855
t.test(1:10, y = c(7:20, 200)) # P = .1245    -- NOT significant anymore
hsu.t.test(1:10, y = c(7:20))
hsu.t.test(1:10, y = c(7:20, 200))

## Traditional interface
with(sleep, t.test(extra[group == 1], extra[group == 2]))
with(sleep, hsu.t.test(extra[group == 1], extra[group == 2]))
## Formula interface
t.test(extra ~ group, data = sleep)
hsu.t.test(extra ~ group, data = sleep)
```

## Bootstrap t-Test
One and two sample bootstrap t-tests with equal or unequal variances in the
two sample case [@Efron1993].
```{r}
boot.t.test(1:10, y = c(7:20)) # without bootstrap: P = .00001855
boot.t.test(1:10, y = c(7:20, 200)) # without bootstrap: P = .1245

## Traditional interface
with(sleep, boot.t.test(extra[group == 1], extra[group == 2]))
## Formula interface
boot.t.test(extra ~ group, data = sleep)
```

## Permutation t-Test
One and two sample permutation t-tests with equal [@Efron1993] or unequal 
variances [@Janssen1997] in the two sample case.
```{r}
perm.t.test(1:10, y = c(7:20)) # without permutation: P = .00001855
## permutation confidence interval sensitive to outlier!
perm.t.test(1:10, y = c(7:20, 200)) # without permutation: P = .1245

## Traditional interface
with(sleep, perm.t.test(extra[group == 1], extra[group == 2]))
## Formula interface
res <- perm.t.test(extra ~ group, data = sleep)
res
```

In case of skewed distributions, one may use function p2ses to compute an
alternative standardized effect size (SES) as proposed by Botta-Dukat [@BottaDukat2018].
```{r}
p2ses(res$p.value)
```


## Multiple Imputation t-Test
Function mi.t.test can be used to compute a multiple imputation t-test by applying
the approch of Rubin [@Rubin1987] in combination with the adjustment of 
Barnard and Rubin [@Barnard1999].
```{r}
## Generate some data
set.seed(123)
x <- rnorm(25, mean = 1)
x[sample(1:25, 5)] <- NA
y <- rnorm(20, mean = -1)
y[sample(1:20, 4)] <- NA
pair <- c(rnorm(25, mean = 1), rnorm(20, mean = -1))
g <- factor(c(rep("yes", 25), rep("no", 20)))
D <- data.frame(ID = 1:45, variable = c(x, y), pair = pair, group = g)

## Use Amelia to impute missing values
library(Amelia)
res <- amelia(D, m = 10, p2s = 0, idvars = "ID", noms = "group")

## Per protocol analysis (Welch two-sample t-test)
t.test(variable ~ group, data = D)
## Intention to treat analysis (Multiple Imputation Welch two-sample t-test)
mi.t.test(res$imputations, x = "variable", y = "group")

## Per protocol analysis (Two-sample t-test)
t.test(variable ~ group, data = D, var.equal = TRUE)
## Intention to treat analysis (Multiple Imputation two-sample t-test)
mi.t.test(res$imputations, x = "variable", y = "group", var.equal = TRUE)

## Specifying alternatives
mi.t.test(res$imputations, x = "variable", y = "group", alternative = "less")
mi.t.test(res$imputations, x = "variable", y = "group", alternative = "greater")

## One sample test
t.test(D$variable[D$group == "yes"])
mi.t.test(res$imputations, x = "variable", subset = D$group == "yes")
mi.t.test(res$imputations, x = "variable", mu = -1, subset = D$group == "yes",
          alternative = "less")
mi.t.test(res$imputations, x = "variable", mu = -1, subset = D$group == "yes",
          alternative = "greater")

## paired test
t.test(D$variable, D$pair, paired = TRUE)
mi.t.test(res$imputations, x = "variable", y = "pair", paired = TRUE)
```

Function mi.t.test also works with package mice by applying function mids2datlist
of package miceadds.
```{r}
library(mice)
library(miceadds)
res.mice <- mice(D, m = 10, print = FALSE)
res.list <- mids2datlist(res.mice)
mi.t.test(res.list, x = "variable", y = "group")
```


## Repeated Measures One-Way ANOVA
We provide a simple wrapper function that allows to compute a classical
repeated measures one-way ANOVA as well as a respective mixed effects model.
In addition, the non-parametric Friedman and Quade tests can be computed.
```{r}
set.seed(123)
outcome <- c(rnorm(10), rnorm(10, mean = 1.5), rnorm(10, mean = 1))
timepoints <- factor(rep(1:3, each = 10))
patients <- factor(rep(1:10, times = 3))
rm.oneway.test(outcome, timepoints, patients)
rm.oneway.test(outcome, timepoints, patients, method = "lme")
rm.oneway.test(outcome, timepoints, patients, method = "friedman")
rm.oneway.test(outcome, timepoints, patients, method = "quade")
```


## Volcano Plots
Volcano plots can be used to visualize the results when many tests have been
applied. They show a measure of effect size in combination with (adjusted)
p values.
```{r, fig.width=7, fig.height=7}
## Generate some data
x <- matrix(rnorm(1000, mean = 10), nrow = 10)
g1 <- rep("control", 10)
y1 <- matrix(rnorm(500, mean = 11.75), nrow = 10)
y2 <- matrix(rnorm(500, mean = 9.75, sd = 3), nrow = 10)
g2 <- rep("treatment", 10)
group <- factor(c(g1, g2))
Data <- rbind(x, cbind(y1, y2))
## compute Hsu t-test
pvals <- apply(Data, 2, function(x, group) hsu.t.test(x ~ group)$p.value,
               group = group)
## compute log-fold change
logfc <- function(x, group){
  res <- tapply(x, group, mean)
  log2(res[1]/res[2])
}
lfcs <- apply(Data, 2, logfc, group = group)
volcano(lfcs, p.adjust(pvals, method = "fdr"), 
        effect.low = -0.25, effect.high = 0.25, 
        xlab = "log-fold change", ylab = "-log10(adj. p value)")
```


## Imputation of Standard Deviations for Changes from Baseline
The function imputeSD can be used to impute standard deviations for changes
from baseline adopting the approach of the Cochrane handbook 
[@Cochrane2019, Section 6.5.2.8].
```{r}
SD1 <- c(0.149, 0.022, 0.036, 0.085, 0.125, NA, 0.139, 0.124, 0.038)
SD2 <- c(NA, 0.039, 0.038, 0.087, 0.125, NA, 0.135, 0.126, 0.038)
SDchange <- c(NA, NA, NA, 0.026, 0.058, NA, NA, NA, NA)
imputeSD(SD1, SD2, SDchange)
```

Correlations can also be provided via argument corr. This option may particularly 
be useful, if no complete data is available.
```{r}
SDchange2 <- rep(NA, 9)
imputeSD(SD1, SD2, SDchange2, corr = c(0.85, 0.9, 0.95))
```


## Pairwise Comparisons
Function pairwise.fun enables the application of arbitrary functions
for pairwise comparisons.
```{r}
pairwise.wilcox.test(airquality$Ozone, airquality$Month, 
                     p.adjust.method = "none")
## To avoid the warnings
library(exactRankTests)
pairwise.fun(airquality$Ozone, airquality$Month, 
             fun = function(x, y) wilcox.exact(x, y)$p.value)
```


## sessionInfo
```{r}
sessionInfo()
```


## References